A Formal Logic for Formal Category Theory

### 2022-10-14 paper/paper.tex
typos in intro

Category theory is a branch of mathematics that studies
higher-dimensional typed algebraic structures.
%
Originally developed for applications to homological algebra, it was
quickly discovered that categorical structures were common in logic and
computer science.  Formal systems like logics, type theories and
programming languages typically have sound and complete models given by
notions of structured categories,\cite{Lawvere69,LambekScott,moggi}.
%
This Curry-Howard-Lambek correspondence~
applies to simply typed lambda calculus~\cite{LambekScott},
computational lambda calculus~\cite{moggi}, linear logic~\cite{girard}
dependent type theory\cite{cartmell1986}, and many other type theories
designed based on category-theoretic semantics.
%
The syntax of a type theory should present an initial object in its
category of models, a category-theoretic reformulation of logical
soundness and completeness.

While this research program has been quite successful,
category-theoretic techniques in computer science are sometimes
criticized for being overly complex and hard to learn, especially for
computer scientists who do not have a strong background in pure
mathematics.
%
While in some ways this could be remedied by better introductory
materials, in some ways category theory is objectively complex.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and limits produce a proliferation of ``naturality'' and
``functoriality'' side-conditions that must be discharged.
%
For example, when constructing an adjoint pair of
functors between two categories, a na\"ive approach would define all of
the data of the action on objects, action on arrows, prove the
functoriality of such actions, as well as construct two families of
transformations, prove they are natural and then finally proving a pair
of equalities relating compositions of natural transformations.
%
Carrying out these proofs explicitly is quite tedious and many newcomers
to are left with the impression that category theory is full of long,
but ultimately trivial constructions.
%
This complexity is compounded when we move from ordinary category theory to
enriched and internal category theory, where constructions must be
additionally proven continuous, monotone, etc, in addition to natural
or functorial.
%
Moreover, these generalizations are often what is needed for programming
language applications; for example, domain- and metric- and
step-index-enriched categories have been used to model recursive
programming languages and internal categories have been used to model
parametricity and gradual
typing\cite{wand197913,topos-of-trees,reflexive-graphs,double-cats-gradual-typing}.

Fortunately, the tools of category theory itself can be employed to
simplify this complexity, specifically the tools of \emph{higher}
category theory.
%
As an analogy in differential calculus, when an adept analyst writes
down a function, they do not expand out the $\epsilon\mathord{-}\delta$
definition of continuity for a function and proceed from first
principles, but rather use certain \emph{syntactic principles} for
defining functions that are continuous by construction --- e.g. that
composition of continuous functions is continuous and that previously
defined functions are known to be continuous.
%
Similar principles apply to category theory itself: functors and
natural transformations are closed under composition and whiskering
operations, and experienced category theorists rely on these syntactic
principles to eliminate the tedium of explicit proofs.
%
In the case of category theory, these principles can be formalized
using algebraic structures such as 2-categories, bicategories,
(virtual) double categories and pro-arrow equipments\cite{bicategories,proarrow-equipments,LEINSTER2002391,Cruttwell2010}, an
approach known as \emph{formal category theory}.
%
In these structures, rather than defining notions of category, functor
and natural transformation from first principles, they are axiomatized
in a manner similar to how a category axiomatizes a notion of space
and homomorphism.
%
The formal approaches also apply to enriched and internal settings, so
\emph{formal} category theory proofs apply to these settings.
%
A downside is that these algebraic structures are quite complicated, and
practitioners typically employ either an algebraic combinator syntax or
a 2-dimensional diagrammatic language that can be quite beautiful and
elegant, but is also somewhat removed from the traditional formulation
of category theory in terms of sets and functions.

In this work, we apply the techniques of categorical logic to define a
more familiar logical syntax for carrying out constructions and proofs
in formal category theory.  
%
We call the resulting theory \emph{virtual equipment type theory} (VETT)
as (hyperdoctrines of) \emph{virtual
  equipments}~\cite{LEINSTER2002391,Cruttwell2010}, a particular semantic model of
formal category theory, provide a sound and complete notion of model for
the theory.
%
VETT provides syntax for categories, functors, profunctors, and natural
transformations, which are defined using familiar term syntax and
$\beta\eta$ reasoning principles for $\lambda$-functions, bound
variables, tuples, etc.  
%
By adhering to a \emph{syntactic discipline}, the logic guarantees
that all functor terms are automatically functorial, all and natural
transformation terms are natural.
%
More specifically, the syntax for transformations is a kind of
\emph{indexed, ordered linear} $lambda$ calculus, where the indexing
ensures that transformations are correctly natural and the ordering and
linearity ensure that the proofs are valid in a large class of enriched
and internal categories, such as enrichment in a non-symmetric monoidal
category.
%
VETT provides an alternative alternative to algebraic and
string-diagram syntaxes for working with virtual equipments,
similar to how the lambda calculus provides an alternative to
categorical combinators and string diagram calculi for cartesian
closed categories.

@@

The syntax of \vett{} is an indexed, ordered linear, proof-relevant
variant of predicate logic over a unary type theory.
%
Just as a predicate logic has a notion of type, term, relation and
implication, \vett{} is based on four analogous category theoretic
concepts: categories, functors, profunctors and natural transformations
of profunctors.
%
Categories are treated like types, and the unary functors we consider in
this paper are each represented by a term whose type is a category and
whose one free variable ranges over a category.
%
The analogue of a relation is a \emph{profunctor} (defined below), which
are written as a kind of set with free category variables.  Like the
restriction to unary functors, we restrict to profunctors with two free
variables.
%
The logic is proof-relevant in that the implications of relations are
generalized to natural transformations of profunctors, and we use a
$\lambda$-calculus notation to describe these ``proof terms''.
%
We can strengthen this analogy into a formal construction: any
construction in \vett{} can be erased to a corresponding construction
or proof in predicate logic, as sets, functions and relations define a 
(somewhat degenerate) virtual equipment.
%% %
%% Our syntax highlights this analogy in that some of our constructions
%% are generalizations of set comprehension or power-sets and we use
%% similar notation.

While the restricted syntax developed in this paper does not express
important concepts such as functor categories or opposite categories,
the restriction is natural in that it corresponds exactly to a
well-understood notion of model, virtual equipments, which can express a
great deal of fundamental results and constructions in category
theory~\cite{riehl_verity_2022,Shulman13}.
%
Moreover, we can work around these unary/binary restrictions to some
extent by viewing the type theory as a domain-specific language embedded
in a metalanguage.  For example, while we cannot talk about functor
categories, we can state a theorem that quantifies over functors using
the meta-language's ``external'' universal quantifier (which does not
have automatic functoriality/naturality properties).  To support this,
\vett includes a third layer, an extensional dependent type theory in
the style of Martin-L\"of type theory. All of our ordered predicate
logic judgments are also be indexed by a context from this dependent
type theory, and the type theory includes universe types for categories,
functors, profunctors and natural transformations.  This allow us to
formalize theorems the logic is too restrictive to encode, analogous to
2-level~\cite{voevodsky13hts,altenkirch+16strict} or indexed type
theories~\cite{isaev21indexed,cervesatopfenning02llf,vakar15linear,krishnaswami+15linear}.

@@

While we emphasize the applications to enriched and internal category
theory in this work, there is potential for more direct application to
programming language semantics.
%
Ordinary predicate logic is the foundation for logical presentations of
logical relations, such as Abadi-Plotkin logic for parametricity and
LSLR and Iris for step-indexed logical relations proofs
\cite{abadi-plotkin,lslr,iris}.
%
We conjecture that \vett{} might similarly serve as the foundation for
a logic of \emph{ordered} structures, which abound in applications:
rewriting and approximation relations can both be modeled as orderings
and logical relations involving these structures are proven to respect
orderings: operational logical relations must be downward-closed and
approximation relations should satisfy transitivity.
%
Just as LSLR and Iris release the user from the syntactic burden of
explicit step-indexing, \vett{} may be used to release the user from
the syntactic burden of proving downward-closure or transitivity
side-conditions.
%
Additionally, \vett{} may serve as the basis of a future domain specific
proof assistant for category theoretic proofs.  To pilot-test this, we
have formalized the syntax of \vett{} in Agda 2.6.2.2, using the rewrite
mechanism to make \vett's substitution and $\beta$-reduction rules
definitional
equalities.\footnote{\url{https://github.com/maxsnew/virtual-equipments/blob/master/agda/STC.agda}}
We have used this lightweight implementation to check a number of
examples.

\textbf{Basics of Profunctors.}  While we assume the reader has some
background knowledge of category theory, we briefly define profunctors,
which are not included in many introductory texts.  Recall that a
category $\cat C$ has a collection of objects and morphisms with
identity and composition, and a functor $F : \cat C \to \cat D$ is a
function on objects and a function on morphisms that preserves identity
and composition.  A category can be thought of as a generalization of a
preordered set, which has a set of elements and a binary \emph{relation}
on its objects satisfying reflexivity and transitivity.  A category is
then a \emph{proof-relevant preorder}, where morphisms are the proofs of
ordering, and the reflexivity and transitivity proofs must identity and
unit equations. A functor is then a \emph{proof-relevant monotone
  function}.  Given categories $\mathcal C$ and $\mathcal D$, a
profunctor $R$ from $\mathcal C$ to $\mathcal D$, written $R : \cat C
\pto \cat D$ is a functor $R : {\cat C}^{op} \times {\cat D} \to \Set$.
Because a profunctor outputs a $\Set$ rather than a proposition, it is
itself a \emph{proof-relevant relation}.  Thinking of categories as
proof-relevant preorders, functoriality says that the profunctor is
downward-closed in $\cat C$ and upward-closed in $\cat D$.  Given
profunctors $R, S : \cat C \pto \cat D$, a homomorphism from $R$ to $S$
is simply a natural transformation, which in the preordered setting is
simply an implication of relations.

@@

%% orphaned --- move later? 
%% For any profunctor $R : \cat C \pto \cat D$ and functors $F : \cat C'
%% \to \cat C$ and $G : \cat D' \to \cat D$, by composition we can define a
%% profunctor $R(F,G)$, the ``restriction'' of $R$ along $F$ and $G$.
The remainder of this paper is organized as follows.  
In Section~\ref{sec:syntax} we introduce the syntax of \vett{}.  In
Section~\ref{sec:examples} we demonstrate how to use our syntax for formal
category theory.  In Section~\ref{sec:semantics}, we develop some model theory
for \vett{}, including a sound and complete notion of categorical model
and sound interpretation in virtual equipments modeling ordinary,
enriched and internal category theory.  In Section~\ref{sec:discussion}, we
discuss related type theories and potential extensions.

>>>>

Category theory is a branch of mathematics that studies
higher-dimensional typed algebraic structures.
%
Originally developed for applications to homological algebra, it was
quickly discovered that categorical structures were common in logic and
computer science.  Formal systems like logics, type theories and
programming languages typically have sound and complete models given by
notions of structured categories~\cite{Lawvere69,LambekScott,moggi}.
%
This Curry-Howard-Lambek correspondence~
applies to simply typed lambda calculus~\cite{LambekScott},
computational lambda calculus~\cite{moggi}, linear logic~\cite{girard}
dependent type theory~\cite{cartmell1986}, and many other type theories
designed based on category-theoretic semantics.
%
The syntax of a type theory should present an initial object in its
category of models, a category-theoretic reformulation of logical
soundness and completeness.

While this research program has been quite successful,
category-theoretic techniques in computer science are sometimes
criticized for being overly complex and hard to learn, especially for
computer scientists who do not have a strong background in pure
mathematics.
%
While in some ways this could be remedied by better introductory
materials, in some ways category theory is objectively complex.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and limits produce a proliferation of ``naturality'' and
``functoriality'' side-conditions that must be discharged.
%
For example, when constructing an adjoint pair of
functors between two categories, a na\"ive approach would define all of
the data of the action on objects, action on arrows, prove the
functoriality of such actions, as well as construct two families of
transformations, prove they are natural and then finally proving a pair
of equalities relating compositions of natural transformations.
%
Carrying out these proofs explicitly is quite tedious and many newcomers
to are left with the impression that category theory is full of long,
but ultimately trivial constructions.
%
This complexity is compounded when we move from ordinary category theory to
enriched and internal category theory, where constructions must be
additionally proven continuous, monotone, etc, in addition to natural
or functorial.
%
However, these generalizations are often exactly what is needed for programming
language applications; for example, domain- and metric- and
step-index-enriched categories have been used to model recursive
programming languages and internal categories have been used to model
parametricity and gradual
typing~\cite{wand197913,topos-of-trees,reflexive-graphs,double-cats-gradual-typing}.

Fortunately, the tools of category theory itself can be employed to
simplify this complexity, specifically the tools of \emph{higher}
category theory.
%
As an analogy in differential calculus, when an adept analyst writes
down a function, they do not expand out the $\epsilon\mathord{-}\delta$
definition of continuity for a function and proceed from first
principles, but rather use certain \emph{syntactic principles} for
defining functions that are continuous by construction --- e.g. that
composition of continuous functions is continuous and that previously
defined functions are known to be continuous.
%
Similar principles apply to category theory itself: functors and
natural transformations are closed under composition and whiskering
operations, and experienced category theorists rely on these syntactic
principles to eliminate the tedium of explicit proofs.
%
In the case of category theory, these principles can be formalized
using algebraic structures such as 2-categories, bicategories,
(virtual) double categories and pro-arrow equipments~\cite{bicategories,proarrow-equipments,LEINSTER2002391,Cruttwell2010}, an
approach known as \emph{formal category theory}.
%
In these structures, rather than defining notions of category, functor
and natural transformation from first principles, they are axiomatized
in a manner similar to how a category axiomatizes a notion of space
and homomorphism.
%
Proofs in formal category theory apply to enriched and internal
settings, which are instances of the formal axioms.
%
A downside is that these algebraic structures are quite complicated, and
practitioners typically employ either an algebraic combinator syntax or
a 2-dimensional diagrammatic language that can be quite beautiful and
elegant, but is also somewhat removed from the traditional formulation
of category theory in terms of sets and functions.

In this work, we apply the techniques of categorical logic to define a
more familiar logical syntax for carrying out constructions and proofs
in formal category theory.  
%
We call the resulting theory \emph{virtual equipment type theory} (VETT)
as (hyperdoctrines of) \emph{virtual
  equipments}~\cite{LEINSTER2002391,Cruttwell2010}, a particular semantic model of
formal category theory, provide a sound and complete notion of model for
the theory.
%
VETT provides syntax for categories, functors, profunctors, and natural
transformations, which are defined using familiar term syntax and
$\beta\eta$ reasoning principles for $\lambda$-functions, bound
variables, tuples, etc.  
%
By adhering to a \emph{syntactic discipline}, the logic guarantees
that all functor terms are automatically functorial, and all natural
transformation terms are natural.
%
More specifically, the syntax for transformations is a kind of
\emph{indexed, ordered linear} $lambda$ calculus, where the indexing
ensures that transformations are correctly natural and the ordering and
linearity ensure that the proofs are valid in a large class of enriched
and internal categories, such as enrichment in a non-symmetric monoidal
category.
%
VETT provides an alternative to algebraic and
string-diagram syntaxes for working with virtual equipments,
similar to how the lambda calculus provides an alternative to
categorical combinators and string diagram calculi for cartesian
closed categories.

@@

The syntax of \vett{} is an indexed, ordered linear, proof-relevant
variant of predicate logic over a unary type theory.
%
Just as a predicate logic has a notion of type, term, relation and
implication, \vett{} is based on four analogous category theoretic
concepts: categories, functors, profunctors and natural transformations
of profunctors.
%
Categories are treated like types, and the unary functors we consider in
this paper are each represented by a term whose type is a category and
whose one free variable ranges over a category.
%
The analogue of a relation is a \emph{profunctor} (defined below), which
is written like a set with free category variables.  Like the
restriction to unary functors, we restrict to profunctors with two free
variables.
%
The logic is proof-relevant in that the implications of relations are
generalized to natural transformations of profunctors, and we use a
$\lambda$-calculus notation to describe these ``proof terms''.
%
This analogy to predicate logic can be made formal: any construction in
\vett{} can be erased to a corresponding construction or proof in
predicate logic, as sets, functions, relations, and implication of
relations define a (somewhat degenerate) virtual equipment.
%% %
%% Our syntax highlights this analogy in that some of our constructions
%% are generalizations of set comprehension or power-sets and we use
%% similar notation.

While the restricted syntax developed in this paper does not express
some important concepts such as functor categories or opposite
categories, the restriction is natural in that it corresponds exactly to
virtual equipments, a well-understood notion of model that can express a
great deal of fundamental results and constructions in category
theory~\cite{riehl_verity_2022,Shulman13}.
%
Moreover, we can work around these unary/binary restrictions to some
extent by viewing the type theory as a domain-specific language embedded
in a metalanguage.  For example, while we cannot talk about functor
categories, we can state a theorem that quantifies over functors using
the meta-language's ``external'' universal quantifier (which does not
have automatic functoriality/naturality properties).  To support this,
\vett{} includes a third layer, an extensional dependent type theory in
the style of Martin-L\"of type theory. All of our ordered predicate
logic judgments are also indexed by a context from this dependent
type theory, and the type theory includes universe types for categories,
functors, profunctors and natural transformations.  This allow us to
formalize theorems the object logic is too restrictive to encode, analogous to
2-level~\cite{voevodsky13hts,altenkirch+16strict} or indexed type
theories~\cite{isaev21indexed,cervesatopfenning02llf,vakar15linear,krishnaswami+15linear}.

@@

While we emphasize the applications to enriched and internal category
theory in this work, there is potential for more direct application to
programming language semantics.
%
Ordinary predicate logic is the foundation for proof-theoretic presentations of
logical relations, such as Abadi-Plotkin logic for parametricity and
LSLR and Iris for step-indexed logical relations proofs
\cite{abadi-plotkin,lslr,iris}.
%
We conjecture that \vett{} might similarly serve as the foundation for
a logic of \emph{ordered} structures, which abound in applications:
rewriting and approximation relations can both be modeled as orderings
and logical relations involving these structures are proven to respect
orderings; operational logical relations must be downward-closed and
approximation relations should satisfy transitivity.
%
Just as LSLR and Iris release the user from the syntactic burden of
explicit step-indexing, \vett{} may be used to release the user from
the syntactic burden of proving downward-closure or transitivity
side-conditions.
%
Additionally, \vett{} may serve as the basis of a future domain specific
proof assistant for category theoretic proofs.  To pilot-test this, we
have formalized the syntax of \vett{} in Agda 2.6.2.2, using the rewrite
mechanism to make \vett's substitution and $\beta$-reduction rules
definitional
equalities.\footnote{\url{https://github.com/maxsnew/virtual-equipments/blob/master/agda/STC.agda}}
We have used this lightweight implementation to check a number of
examples.

\textbf{Basics of Profunctors.}  While we assume the reader has some
background knowledge of category theory, we briefly define profunctors,
which are not included in many introductory texts.  Recall that a
category $\cat C$ has a collection of objects and morphisms with
identity and composition, and a functor $F : \cat C \to \cat D$ is a
function on objects and a function on morphisms that preserves identity
and composition.  A category can be thought of as a generalization of a
preordered set, which has a set of elements and a binary \emph{relation}
on its objects satisfying reflexivity and transitivity.  A category is
then a \emph{proof-relevant preorder}, where morphisms are the proofs of
ordering, and the reflexivity and transitivity proofs must satisfy identity and
unit equations. A functor is then a \emph{proof-relevant monotone
  function}.  Given categories $\mathcal C$ and $\mathcal D$, a
profunctor $R$ from $\mathcal C$ to $\mathcal D$, written $R : \cat C
\pto \cat D$ is a functor $R : {\cat C}^{op} \times {\cat D} \to \Set$.
Because a profunctor outputs a $\Set$ rather than a proposition, it is
itself a \emph{proof-relevant relation}.  Thinking of categories as
proof-relevant preorders, functoriality says that the profunctor is
downward-closed in $\cat C$ and upward-closed in $\cat D$.  Given
profunctors $R, S : \cat C \pto \cat D$, a homomorphism from $R$ to $S$
is a natural transformation, which in the preordered setting is
simply an implication of relations.

@@

%% orphaned --- move later? 
%% For any profunctor $R : \cat C \pto \cat D$ and functors $F : \cat C'
%% \to \cat C$ and $G : \cat D' \to \cat D$, by composition we can define a
%% profunctor $R(F,G)$, the ``restriction'' of $R$ along $F$ and $G$.
\textbf{Outline.}  In Section~\ref{sec:syntax} we introduce the syntax
of \vett{}.  In Section~\ref{sec:examples} we demonstrate how to use our
syntax for formal category theory.  In Section~\ref{sec:semantics}, we
develop some model theory for \vett{}, including a sound and complete
notion of categorical model and sound interpretation in virtual
equipments modeling ordinary, enriched and internal category theory.  In
Section~\ref{sec:discussion}, we discuss related type theories and
potential extensions.
